Optimally Efficient Lock-free Collaborative Text Editing

### 2013-01-12 paper-collabed.tex
proofreading corrections

% Comments --
% 12/10/12 -
%  - incorporate previous work on CRDTs by Preguica, Weiss,  Urso, Molli, Oster, etc.
%
% 03/01/10 - 
%   -Benchmark space complexity of Node extension vs node of size one
%   -Explain the three types of linked lists (section 6)
%   -Work the clash example with the hashtable for node ids, linked list of
%    of global subnodes, linked list of local subnodes.
%   -Create git repository for Dissertation and Collabed.
%   -
% 02/23/10 - Work on the optimization section and lead it into experimental
% results.

@@

\begin{abstract}
In this paper we present a new collaborative text editing data type,  
the Monotone Shared Edit Tree (MSET). This algorithm is an extension
and optimization of the TreeDoc datatype of Preguica and Shapiro.  The MSET data type is an example of
a Commutative Replicated Data Type. It does not rely on locking, is not based on operational transformation, 
and does not require a central server to serialize the
edit operations.
MSET allows multiple users to locally 
edit their own copies  of a shared text document
by inserting or deleting characters
while simultaneously sending out their local edits 
to remote users and processing edits from remote users 
on their local copy.
Each edit operation (whether generated locally or remotely) 
requires $O(\log(N))$ time to process, where
N is the total number of characters that have inserted 
into the user's document before the operation in
question is applied (including those characters that were later deleted.)
The MSET algorithm is based on 
the Model-View-Controller paradigm in which the
model underlying the text document is a simple and 
natural type of tree that represents the full collaborative edit process.
The edits that are allowed for this tree are monotone in the sense that
the individual edits add a component to the tree which can never be removed
though it can be evolved. For example, deletion is handled by setting the
visibility attribute of the character to ``hidden''. The operations are also
designed to facilitate shared editing in the sense that the order in which
they are applied does not affect the eventual result.
The MSET algorithm is convergent in the sense that 
if all users stop generating
editing operations, then when all operations have 
been received and processed, all users will have
the same document. 
A collaborative editor based on the MSET algorithm 
has been successfully introduced as a plugin into widely used
editors such as JEdit, Eclipse and Netbeans and 
these collaborative editors have been used in classroom situations
with positive feedback from the users (both students and faculty).
\end{abstract}
\newpage
\tableofcontents
\newpage

@@

Most of the early work on collaborative editing used an operational transformation approach with a central server that would serialize all edit operations  and rebroadcast them to the clients \cite{nichols_high-latency_1995}. The server and clients would transform the edit operations they received from each other before applying them to account for differences in the remote documents when the original edit operations were applied. Many extensions and variations of this early work have been developed over the ensuing 25 years (see e.g. \cite{sun_operational_2004}).

@@

Our work, MSET, is a generalization and optimization of the TreeDoc data type  of Preguica, et. al.  \cite{preguica_commutative_2009}. It has been available since 2007 as an open source project hosted at sourceforge \cite{granville_collabed_2007} and has been used as the foundation of the CollabEd system  \cite{granville_collabed:_2009}, which provides a common plugin architecture for code editors allowing developers on different platforms (e.g. jEdit, NetBeans, Eclipse) to collaboratively edit a code file while still using the IDE that they prefer.

@@

\subsection{The Collision Problem}
The fundamental problem in collaborative editing of any documents  
(of any type, not just text)
is dealing with collisions that occur when two 
users (say $A$ and $B$) apply different operators (say $\alpha$ and $\beta$) 
to a shared document. Assuming that both users
start off with the same document $D$, the two users will have two (usually) 
different documents $\alpha(D)$ and $\beta(D)$.
A collaborative editor needs to resolve this conflict, so that both users 
return to the state of having the same document $D'$.
This can be done 
by introducing transformed operations $\alpha'=T_\beta(\alpha)$ and 
$\beta'=T_\alpha(\beta)$ as shown in the diagram below,
where $T$ is a transform operator:
\[
\begin{diagram}
\node{D} \arrow{e,t}{\alpha} \arrow{s,l}{\beta} \node{\alpha(D)} \arrow{s,r}{\beta'=T_\alpha(\beta)} \\
\node{\beta(D)} \arrow{e,t}{\alpha'=T_\beta(\alpha)} \node{D'}
\end{diagram}
\]
and the transform operator must be defined so that this diagram commutes, 
that is, there must be an operator $\gamma$ on
$D$ such that:
\[
\alpha'(\beta(D)) = \beta'(\alpha(D)) = \gamma(D)
\]
If this operation holds, we can let $(\alpha \vert \beta)$ denote the combined operation $\gamma$.

If the operators are invertible then one can define 
the transform operator $T$ in terms of the collision
operator $\gamma$, as follows. Since the operators $\alpha$ 
and $\beta$ are invertible, there must be
operators $\alpha^{-1}$ and $\beta^{-1}$ such that
\[
\alpha^{-1}(\alpha(D)) = D \;\;
\beta^{-1}(\beta(D)) = D 
\]
we then define T as follows:
\[
T_\alpha(\beta) = (\alpha \vert \beta) \circ \alpha^{-1}
\]
Once a collision operator is defined, 
one can implement a collaborative editor which uses a central
server to serialize all of the edit operations into a single stream 
and which applies the transforms to resolve
conflicts among users.

@@

@@

\subsubsection{Initial state}
All users start out with the empty string $\S^u_i=""$ where the
tree representation $T^u_i$ is a single node owned by user $u_0$, the superuser,
and containing only the start and end marker symbols $<^0_0$ and $>^0_0$ for the 0th node of user 0. These markers appear only in 
 the edit-string view.
\begin{align*}
E^u_0 &=& <_0^0 \;\;\;\;\;\; >_0^0 \\
R^u_0 &=& \epsilon\\
S^u_0 &=& \epsilon
\end{align*}
where $\epsilon$ denotes the empty string.
MSET only allows users to edit the standard view, $S^u_i$, of their document. They may insert or delete characters anywhere in $S^u_i$, but the other two views are purely read-only.

@@

\subsection{Creating conflicts in collaborative editing}
Next, suppose user $u3$ 
simultaneously inserts the adjective "illustrative "
before the word "example" in their own
local copies and that both users broadcast their edits to their peers, which
includes user 2 who is just watching the other two edit.  
In a high latency environment (or in a setting where the users are queueing 
their edits before sending them out), user 2 would create its own tree
but with a different node attached at position 11 of the $u1/n0$. This tree
is shown in the right side of Figure \ref{fig:tree11}. Again, if user 3
is typing the word character by character then the first character will generate
the node $u3/n0$ containing only one character.

@@

In the case where the new nodes were created one character at a time, rather
than using cut/paste to insert the words, the conflict would occur with two
nodes that each contain a single character and will be resolved in the same
way.  Also, the users doing the editing $u1$ and $u3$ will receiving the
conflicting edit operation and will create the correspond nodes and modify
their string views without changing the relative cursor position. Thus as they
continue typing their edits will be used to extend their nodes and the
same tree will result as if they had used cut/paste to insert the words.

@@

\subsection{Inserting into a tree with hidden characters}
Suppose now that user 2 has received all of these edits
and wishes to insert the word ``easy''
into the place where ``interested'' had just been deleted by user 1.
This would result in the following string:
\[
\text{This is an easy illustrative example.}
\]
The point we want to make here is that there are several places that the new node
$X = <^2_0 \text{easy} >^2_0$ could be inserted in the tree
which would place it between "an" and "illustrative" in the string view.
Several possibilities are shown below and their corresponding edit-trees
are snown in Figures \ref{fig:tree14a} and \ref{fig:tree14b}:
\begin{itemize}
\item {\bf leftmost insertion}
The first possibility is to insert the new node directly before the
first character (visible or hidden) after the insertion point. This
is in fact what the current implementation of MSET would select
for this particular example,
but any of these operations could be used.
\[
 <^0_0 <^1_0 
 \text{This is an} 
   <^1_1 <^2_0 \text{easy} >^2_0 \text{\sout{interesting}} >^1_1
  <^3_0 \text{illustrative} >^3_0
  \text{example.} >^1_0 >^0_0
\]
\item {\bf rightmost insertion}
The next is to insert it as far to the right as possible in
the edit-tree.
\[
 <^0_0 \;<^1_0 
 \text{This is an} 
   \;<^1_1 \text{\sout{interesting}} >^1_1\;
  <^3_0 \;<^2_0 \text{easy} >^2_0 \;\text{illustrative} >^3_0\;
  \text{example.} >^1_0\; >^0_0
\]
\item {\bf insertion in a hidden string}
Yet another possibility is to insert it somewhere else (anywhere) in the node $u1/n1$
which contains only hidden text.
\[
 <^0_0 <^1_0 
 \text{This is an} 
   <^1_1 \text{\sout{interesting}} <^2_0 \text{easy} >^2_0>^1_1
  <^3_0 \text{illustrative} >^3_0
  \text{example.} >^1_0 >^0_0
\]
\item {\bf insertion as a child of $u1/n0$}
The final possibility is to insert it as another node at the same level
as $u1/n1$ and $u3/n0$, which would result in
\[
 <^0_0 <^1_0 
 \text{This is an} 
   <^1_1 \text{\sout{interesting}}>^1_1
   <^2_0 \text{easy} >^2_0
   <^3_0 \text{illustrative} >^3_0
 \text{example.} >^1_0 >^0_0
\]
\end{itemize}
These are all the possible ways in which the edit observed in the string view
could be obtained by a valid edit on the edit-tree.

@@

\subsection{General features revealed by the example}
This extended example demonstrates the fundamental ideas behind the MSET
approach. Each user $u_i$ represents the shared document as an edit-tree $T_i$
which has two different views: the standard view $\alpha_i = f(T_i)$ which
is the string of visible characters in $T_i$, and an edit-string view
$\beta_i = g(T_i)$ which contains the hidden characters as well as the
start and end markers for the nodes. The user can insert and delete anywhere
in the standard view, but can only mark characters as "hidden" in the
edit-string view and can only insert at selected places in the edit-string
view. The string operations are converted to edit-tree operations which are
broadcast to all of the peers in the collaborative editing group. Those users
apply the remote operations they receive in any order they choose, provided only
that they wait for the target node of an operation to be created before the'
operation is applied.

We will describe the algorithm in more detail below and we will show that it
can be implemented in such a way that each local and remote edit operation
can be performed in time $O(k\log(N))$ where $k$ is the number of characters
in the operation and $N$ is the size of the edit-string.

@@

As mentioned in the introduction the MSET model is based on the notion that 
each user maintains an edit tree as the underlying model of the textarea in
which they are editing. In this section, we describe an algorithm for 
collaboratively editing the edit-trees directly. In a later section we will show
how to extend this to an editor of edit-strings as well as strings in
the standard view. Then we'll show how to implement these algorithms
efficiently. In this section, we assume that the users are only interested in editing
the trees directly.

@@

\subsection{Valid edit operations on an edit-tree}
We place restrictions on the type of edits that can be performed
on an edit tree. Indeed, there are only three actions a user $u$ can take
when editing their copy of the shared edit tree. We will indicate nodes 
of the edit-tree being edited by
the notation $u/n$ where $u$ is the user who created the node and $n$ is an
integer uniquely identifying this node among all nodes created by the user.

@@

Let $u$ be the user performing one of the following tree tree edit operations
on the target node {\tt v/m:p} to create a new node {\tt u/n:q}.
\begin{itemize}
\item ${\tt treeextend}(v/m:p,c)\;\;\text{if u=v}$ \newline
First, user $u$ can extend the text in the node $v/m$
provided it is the owner (that is $v=u$). It extends the node by appending
some character $c$ to the location $p$ at the end of the node,
Note however, that users can not modify
the text in any node that they do not own and can not change any of the text
once it has been put into the node.
\item ${\tt treeinsert}(v/m:p,q,u/n:1,c) \;\; \text{if $q\le p$}$
\newline
Second, user $u$ can insert a new node $u/n$ containing a character $c$
at any position $q\le p$ of the node, 
$v/m$ provided
they have not already inserted a node there. If one or more nodes have already been inserted there, then they must be arranged in order of the userid and the new node is inserted into this sorted sequence by userid.
\item ${\tt treehide}(v/m:p,q)  \;\; \text{if $q\le p$}$\newline
Third, user $u$ can set the visibility of a character at position $q < p$
in  $v/m:p$ to false.
When characters are initially added to a node they have visibilty true, but once
set to false, the visibilty must remain false. 
\end{itemize}
For all of these operations, we say that the node $(v/m:p)$ is the target of the edit operation and the node that is created, either $(v/m:p+1)$ for an extension or $(u/n:1)$ for a new node, is the result.

@@

Given these definitions and rules we see that a group of distributed users can easily edit a 
shared tree by keeping their own representation of the tree and broadcasting
their edits to their peers. Observe that these edit operations commute in
the following sense. The only place where two user's could have a editing
conflict is when they both want to insert a string of characters at a particular
position $q$ in some node $v/m:p$, but in this case their insertions are always
ordered by userid and this resolves the conflict. No matter which order the
operations are applied the same tree results. This is phrased more formally
in the following lemma:

@@

\begin{proof}
First, the sequence $S_{i}=\sigma_i(S_{i-1})$ is well-defined because we have
stipulated that the target of each $\sigma_i$ must exist in $S_{i-1}$ so the
operator can be applied. To see that the two resulting trees are equivalent
we simply observe that the extend operations for a node $v/m:p$ must appear
in the same order in the $\sigma_i$ as the $\tau_i$ and since they effect only
the node $v/m$ the node $v/m$ will contain the same string in both $S_n$ and $T_n$.
The insert operations for a given node $v/m$ on the other hand can appear
in any order and their effect will be to add children to positions $q$
of the final node. When multiple insertions attach to the same position, they
will be ordered by user id. Since they are inserted into an ordered list,
the order in which they are inserted doesn't matter.
\end{proof}

@@

The model is that we will have a set $G$ of clients $C=C_1,\ldots,C=C_n$
all starting with the empty tree. Each client $C_i$ has a unique userid $u_i$,
a local copy $T_i$ of the shared edit tree,
a list $Q_i$ of edit operations that other users have applied but which it
has not yet processed, and
and a map $H_i$ from nodeids $v/m$ to nodes of the edit tree or to sets of tree-edit operations using a union type. If the tree $T_i$ contains a node $A$ with nodeid $v/m:p$ then $H_i(v/m:p)=A$, otherwise $H_i(v/m:p)$ is a list $S$ of tree-edit operations taken from $Q_i$ whose target is $v/m:p$.

We assume that the clients observe the following protocol. 
In particular user $u_i$ can do the following:
\begin{itemize}
\item Perform a valid operation $\tau$ on their local copy of the tree to
produce a new (or extended) node $n$ with nodeid $u/n:q$,
after which they set $H_i(u/n:q)$ to $n$ and they also add $\tau$ to $Q_j$ for
all $j\ne i$.
\item Take an operation $\tau$ from $Q_i$. If the target of $\tau$ is
in $T_i$ then $\tau$ is applied to the tree to produce a node $v/m:p$
and the tree-edit operations in $H_i(v/m:p)$, if any, are inserted back into $Q_i$
 and $H_i(v/m:p)$ is set to $n$.
\end{itemize}
Note, this model can easily be extended to allows for late joiners by having users
maintain a history list $Q'_i$ of all operations they have applied to their tree.

Any user $k$ who joins the group notifies all users it has joined and waits for confirmation from all users. Once this is obtained, it 
requests some client $C_j$ to forward its current history list $Q'_k$ and its current queue, 
$Q_k$ into its own edit queue $Q_i$.
This can be an incremental copy at user $j$'s leisure depending
on the workload.  This may require user $k$ to check each new operation
it receives to make sure that it has not already been received and processed,
but this is not too expensive.

@@

\section{From edit-trees to an optimally efficient collaborative editor}
\label{sec:proof}
In this section we will define the MSET data type $M$ which simultaneously maintains
four views of an editing session
\begin{itemize}
\item $M.S$ - the standard string view of a text-editing session, a sequence of characters from a character set $\Sigma$
\item $M.R$ - a revisions view which is a sequence of characters which each have a visibility attribute (visible or hidden), this is similar to the tracking-changes view of modern text editors
\item $M.E$ - an edit-string view which is a sequence of attributed characters and marker symbols (start markers $<^u_n$ and end markers $>^u_n$)
\item $M.T$ - an edit tree
\end{itemize}
The three string views are all views of the edit tree $M.T$ obtained by traversing the tree in infix order and recording the characters and or symbols of the specified type.

@@

 {{\tt delete($k$)}} 
Lets handle the deletion case first as it is easiest. The first step is to lookup the element $e$ at position $k$ in $M.S$. This can be done in time $O(\log(N))$.
We can then use $e$ to lookup the node $n$ and offset $q$ of $e$ in that node.
This takes time $O(1)$ the node and offset are fields of $e$. Finally, we lookup the
nodeid $v/m$ and length $p$ of $n$ and generate the operation $treehide(v/m:p,q)$. This operation takes time $O(\log(N))$. In pseudocode, we have
\begin{verbatim}
delete(k)
  e = M.I.lookupS(k);
  v/m:p = e.nodeid;  q = e.offset
  return treehide(v/m:p,q)
\end{verbatim}

@@

\subsubsection{Case 3a: $g$ is a non-marker element}
In this case, $g$ is either $f$ or a hidden character, but in any case, we can insert $\alpha$ between $e$ and $g$. So in time $O(1)$ we lookup the nodeid $v/m$ and
offset $q$ of $g$ and generate the tree edit operation 
$treeinsert(v/m:p,q,u/n:1,c)$

@@

\begin{verbatim}
treehide(v/m:p,q)
  n = M.lookupNode(v/m:p);
  e = n.elt[q]
  M.I.hide(e)
  /* return view op to update M.S, M.R, M.E */
  return delete(e.listOffset)
\end{verbatim}

>>>>

% Comments --
% 1/11/13 - Proofreading level revision of paper...
%
% 12/10/12 -
%  - incorporate previous work on CRDTs by Preguica, Weiss,  Urso, Molli, Oster, etc.
%
% 03/01/10 - 
%   -Benchmark space complexity of Node extension vs node of size one
%   -Explain the three types of linked lists (section 6)
%   -Work the clash example with the hashtable for node ids, linked list of
%    of global subnodes, linked list of local subnodes.
%   -Create git repository for Dissertation and Collabed.
%   -
% 02/23/10 - Work on the optimization section and lead it into experimental
% results.

@@

\begin{abstract}
In this paper we present a new collaborative text editing data type,  
the Monotone Shared Edit Tree (MSET). This algorithm is an extension
and optimization of the TreeDoc datatype of Preguica and Shapiro 
which is one of several new collaborative editing models which are based
on the concept of a Commutative Replicated Data Type (CRDT). 
These models do not rely on locking, are not based on operational transformation,
maintain a local copy of the document with each client,  
and do not require a central server to serialize the edit operations.
Editors based on CRDT allow multiple users to locally 
edit their own copies  of a shared text document
by inserting or deleting characters
while simultaneously sending out local edits 
to remote users and processing edits from remote users 
on their local copy. The primary innovation with MSET is that it
provides an optimally efficient algorithm where 
each edit operation (whether generated locally or remotely) 
requires $O(\log(N))$ time to process, where
N is the total number of characters that have inserted 
into the user's document before the operation in
question is applied (including those characters that were later deleted).
%The MSET algorithm is based on 
%a Model-View-Controller paradigm in which the
%model underlying the text document is a simple and 
%natural type of tree that represents the full collaborative edit process.
%The edits that are allowed for this tree are monotone in the sense that
%the individual edits add a component to the tree which can never be removed
%though it can be evolved in a monotone fashion. For example, deletion is handled by changing the
%visibility attribute of a character from ``visible" to ``hidden''. The tree operations facilitate shared editing as the order in which
%they are applied does not affect the eventual result.
%The key property of the MSET algorithm is that it is convergent, i.e.  
%if all users stop generating
%editing operations, then when all operations have 
%been received and processed, all users will have
%the same document. 
A collaborative editor, CollabEd, 
based on an optimized version of the MSET algorithm 
has been successfully introduced as a plugin into widely used
editors such as JEdit, Eclipse and Netbeans and 
these collaborative editors have been used in classroom situations
with positive feedback from the users (both students and faculty).
\end{abstract}
\newpage
\tableofcontents
\newpage

@@

Most of the early work on collaborative editing, beginning with dOPT, used an operational transformation approach with a central server that would serialize all edit operations  and rebroadcast them to the clients \cite{nichols_high-latency_1995}. The server and clients would transform the edit operations they received from each other before applying them to account for differences in the remote documents when the original edit operations were applied. Many extensions and variations of this early work have been developed over the ensuing 25 years (see e.g. \cite{sun_operational_2004}).

@@

Our work, MSET, is a generalization and optimization of the TreeDoc data type  of Preguica, et. al.  \cite{preguica_commutative_2009}. It has been available since 2007 as an open source project hosted at sourceforge \cite{granville_collabed_2007} and has been used as the foundation of the CollabEd system  \cite{granville_collabed:_2009}, which provides a common plugin architecture for code editors allowing developers on different platforms (e.g. jEdit, NetBeans, Eclipse) to simultaneously edit the same code file using the IDE that they individually prefer.

@@

\subsection{The Collision Problem}
The fundamental problem in collaborative editing of any document  
(of any type, not just text)
is dealing with collisions that occur when two 
users (say $A$ and $B$) apply different operators (say $\alpha$ and $\beta$) 
to a shared document. Assuming that both users
start off with the same document $D$, the two users will have two (usually) 
different documents $\alpha(D)$ and $\beta(D)$.
A collaborative editor needs to resolve this conflict, so that both users 
return to the state of having the same document $D'$.
In the Operational Transformation approach, this is done 
by introducing transformed operations $\alpha'=T_\beta(\alpha)$ and 
$\beta'=T_\alpha(\beta)$ as shown in the diagram below,
where $T$ is a transform operator:
\[
\begin{diagram}
\node{D} \arrow{e,t}{\alpha} \arrow{s,l}{\beta} \node{\alpha(D)} \arrow{s,r}{\beta'=T_\alpha(\beta)} \\
\node{\beta(D)} \arrow{e,t}{\alpha'=T_\beta(\alpha)} \node{D'}
\end{diagram}
\]
and the transform operator must be defined so that this diagram commutes, 
that is, there must be an operator $\gamma$ on
$D$ such that:
\[
\alpha'(\beta(D)) = \beta'(\alpha(D)) = \gamma(D) = D'
\]
If this operation holds, we can let $(\alpha \vert \beta)$ denote the combined operation $\gamma$.

If the operators are left invertible then one can define 
the transform operator $T$ in terms of the collision
operator $\gamma$, as follows. Since the operators $\alpha$ 
and $\beta$ are invertible, there must be
operators $\alpha^{-1}$ and $\beta^{-1}$ such that
\[
\alpha^{-1}(\alpha(D)) = D \;\;
\beta^{-1}(\beta(D)) = D 
\]
we then define T as follows:
\[
T_\alpha(\beta) = (\alpha \vert \beta) \circ \alpha^{-1}
\]
Once a collision operator is defined, 
one can implement a collaborative editor which uses a central
server to serialize all of the edit operations into a single stream 
and which applies the transforms to resolve
conflicts among users.

@@

A final concern is that if the network traffic is temporarily halted and both
clients perform $n$ operations without processing any of the other client's
operations, then the time for them to resync is $O(n^2)$. To see this, let $D_{i,j}$ denote the document obtained by processing $i$ operations from user 1 and $j$ operations from user 2, where both clients start with $D_{0,0}$. If user 1 is at $D_{i,0}$ and user 2 is at $D_{0,j}$, then they will both need to compute $D_{i,j}$ for all $i,j\in[0,n]$ to be able to transform $D_{0,0}$ into $D_{n,n}$. Typically,
user 1 will calculate $D_{i,1}$ for all $i$ from 1 through $n$, then $D_{i,2}$ for all $i$, etc. while user 2 will calculate $D_{1,j}$ for all $j$, then $D_{2,j}$ etc. The MSET algorithm we propose has no added complexity for such network delays and, in fact, we purposely allow the users to enqueue incoming operations (and/or outgoing operations) and this feature would render an Operational Transform system unusable.

@@

\subsubsection{Initial state}
All users start out with the empty string $S^u_i=""$ where the
tree representation $T^u_i$ is a single node owned by user $u_0$, the superuser,
and containing only the start and end marker symbols $<^0_0$ and $>^0_0$ for the 0th node of user 0. These markers appear only in 
 the edit-string view.
\begin{align*}
E^u_0 &=& <_0^0 \;\;\;\;\;\; >_0^0 \\
R^u_0 &=& \epsilon\\
S^u_0 &=& \epsilon
\end{align*}
where $\epsilon$ denotes the empty string.
MSET only allows users to edit the standard view, $S^u_i$, of their document. They may insert or delete characters anywhere in $S^u_i$, but the other two views are purely read-only.

@@

\subsection{Creating conflicts in collaborative editing}
Next, suppose user $u3$ 
simultaneously inserts the adjective "illustrative "
before the word "example" in her own
local copy and that both users $u1$ and $u3$ broadcast their edits to their peers, which
includes user $u2$ who is just watching the other two edit.  
In a high latency environment (or in a setting where the users are queueing 
their edits before sending them out), user $u2$ would create its own tree
but with a different node attached at position 11 of the $u1/n0$. This tree
is shown in the right side of Figure \ref{fig:tree11}. Again, if user $u3$
is typing the word character by character then the first character will generate
the node $u3/n0$ containing only one character.

@@

In the case where the new nodes were created one character at a time, rather
than using cut/paste to insert the words, the conflict would occur with two
nodes that each contain a single character and will be resolved in the same
way.  Also, the users doing the editing, $u1$ and $u3$, will receive a
conflicting edit operation and will create the correspond nodes and modify
their string views without changing the relative cursor position. Thus as they
continue typing their edits will be used to extend their nodes and the
same tree will result as if they had used cut/paste to insert the words.

@@

\subsection{Inserting into a tree with hidden characters}
Suppose now that user 2 has received all of these edits
and wishes to insert the word ``easy''
into the place where ``interested'' had just been deleted by user 1.
This would result in the following string:
\[
\text{This is an easy illustrative example.}
\]
The point we want to make here is that there are several places that the new node
$X = <^2_0 \text{easy} >^2_0$ could be inserted in the tree
which would place it between "an" and "illustrative" in the string view.
Several possibilities are shown below and their corresponding edit-trees
are snown in Figures \ref{fig:tree14a} and \ref{fig:tree14b}:
\begin{itemize}
\item {\bf leftmost insertion}
The first possibility is to insert the new node directly before the
first character (visible or hidden) after the insertion point. This
is in fact what the current implementation of MSET would select
for this particular example,
but any of these operations could be used.
\[
 <^0_0 <^1_0 
 \text{This is an} 
   <^1_1 <^2_0 \text{easy} >^2_0 \text{\sout{interesting}} >^1_1
  <^3_0 \text{illustrative} >^3_0
  \text{example.} >^1_0 >^0_0
\]
\item {\bf rightmost insertion}
The next is to insert it as far to the right as possible in
the edit-tree.
\[
 <^0_0 \;<^1_0 
 \text{This is an} 
   \;<^1_1 \text{\sout{interesting}} >^1_1\;
  <^3_0 \;<^2_0 \text{easy} >^2_0 \;\text{illustrative} >^3_0\;
  \text{example.} >^1_0\; >^0_0
\]
\item {\bf insertion in a hidden string}
Yet another possibility is to insert it somewhere else (anywhere) in the node $u1/n1$
which contains only hidden text.
\[
 <^0_0 <^1_0 
 \text{This is an} 
   <^1_1 \text{\sout{interesting}} <^2_0 \text{easy} >^2_0>^1_1
  <^3_0 \text{illustrative} >^3_0
  \text{example.} >^1_0 >^0_0
\]
\item {\bf insertion as a child of $u1/n0$}
The final possibility is to insert it as another node at the same level
as $u1/n1$ and $u3/n0$, which only works because coincidentally $u2<u3$. This  would result in
\[
 <^0_0 <^1_0 
 \text{This is an} 
   <^1_1 \text{\sout{interesting}}>^1_1
   <^2_0 \text{easy} >^2_0
   <^3_0 \text{illustrative} >^3_0
 \text{example.} >^1_0 >^0_0
\]
\end{itemize}
These are all the possible ways in which the edit observed in the string view
could be obtained by a valid edit on the edit-tree.

@@

\subsection{General features revealed by the example}
This extended example demonstrates the fundamental ideas behind the MSET
approach. Each user $u_i$ represents the shared document as an edit-tree $T_i$
which has three different views: the standard view  which
is the string of visible characters in $T_i$, a revision string view which shows both visible and hidden characters (possibly with strike-through font) and an edit-string view which contains the hidden characters as well as the
start and end markers for the nodes. The user can insert and delete anywhere
in the standard view, but can only mark characters as "hidden" in the
edit-string view and can only insert at selected places in the edit-string
view. The string operations are converted to edit-tree operations which are
broadcast to all of the peers in the collaborative editing group. Those users
apply the remote operations they receive in any order they choose, provided only
that they wait for the target node of an operation to be created before the'
operation is applied.

We will describe the algorithm in more detail below and we will show that it
can be implemented in such a way that each local and remote single character 
edit operation
can be performed in time $O(\log(N))$ where $N$ is the size of the edit-string.

@@

As mentioned in the introduction the MSET model is based on the notion that 
each user maintains an edit tree as the underlying model of the textarea in
which they are editing. In this section, we describe an algorithm for 
collaboratively editing the edit-trees directly. In a later section we will show 
how to implement these algorithms
efficiently. In this section, we assume that the users are only interested in editing
the trees directly. This algorithm is essentially the same as the TreeDoc data type of Preguica, et. al. The main innovation of this paper is that it shows how to use this tree structured data type to implement an optimally efficient text editor.

@@

\subsection{Valid edit operations on an edit-tree}
We place restrictions on the type of edits that can be performed
on an edit tree.  We will indicate nodes 
of the edit-tree being edited by
the notation $u/n$ where $u$ is the user who created the node and $n$ is an
integer uniquely identifying this node among all nodes created by the user.

@@

Let $u$ be the user performing one of the following tree tree edit operations
on the target node {\tt v/m:p} to create a new node {\tt u/n:q}.
\begin{itemize}
\item ${\tt treeextend}(v/m:p,c)\;\;\text{if u=v}$ \newline
First, user $u$ can extend the text in the node $v/m$
provided it is the owner (that is $v=u$). It extends the node by appending
some character $c$ to the location $p$ at the end of the node,
Note however, that users can not modify
the text in any node that they do not own and can not change any of the text
once it has been put into the node.
\item ${\tt treeinsert}(v/m:p,q,u/n:1,c) \;\; \text{if $q\le p$}$
\newline
Second, user $u$ can insert a new node $u/n$ containing a character $c$
at any position $q\le p$ of the node, 
$v/m$ provided
they have not already inserted a node there. If one or more nodes have already been inserted there, then they must be arranged in order of the userid and the new node is inserted into this sorted sequence by userid.
\item ${\tt treehide}(v/m:p,q)  \;\; \text{if $q < p$}$\newline
Third, user $u$ can set the visibility of a character at position $q < p$
in  $v/m:p$ to false.
When characters are initially added to a node they have visibilty true, but once
set to false, the visibilty must remain false. 
\end{itemize}
For all of these operations, we say that the node $(v/m:p)$ is the target of the edit operation and the node that is created, either $(v/m:p+1)$ for an extension or $(u/n:1)$ for a new node, is the result.

@@

Given these definitions and rules we see that a group of distributed users can easily edit a 
shared tree by keeping their own representation of the tree and broadcasting
their edits to their peers. Observe that these edit operations commute in
the following sense. The only place where two users could have a editing
conflict is when they both want to insert a string of characters at a particular
position $q$ in some node $v/m:p$, but in this case their insertions are always
ordered by userid and this resolves the conflict. No matter which order the
operations are applied the same tree results. This is phrased more formally
in the following lemma:

@@

\begin{proof}
First, the sequence $S_{i}=\tau'_i(S_{i-1})$ is well-defined because we have
stipulated that the target of each $\tau'_i$ must exist in $S_{i-1}$ so the
operator can be applied. To see that the two resulting trees are equivalent
we simply observe that the extend operations for a node $v/m:p$ must appear
in the same order in the $\tau'_i$ as the $\tau_i$ and since they effect only
the node $v/m$ the node $v/m$ will contain the same string in both $S_n$ and $T_n$.
The insert operations for a given node $v/m$ on the other hand can appear
in any order and their effect will be to add children to positions $q$
of the final node. When multiple insertions attach to the same position, they
will be ordered by user id. Since they are inserted into an ordered list,
the order in which they are inserted doesn't matter.
\end{proof}

@@

The model is that we will have a set $G$ of clients $C_1,\ldots,C_n$
all starting with the empty tree. Each client $C_i$ has a unique userid $u_i$,
a local copy $T_i$ of the shared edit tree,
a list $Q_i$ of edit operations that other users have applied but which it
has not yet processed, and
and a map $H_i$ from nodeids $v/m$ to nodes of the edit tree or to sets of tree-edit operations using a union type. If the tree $T_i$ contains a node $A$ with nodeid $v/m:p$ then $H_i(v/m:p)=A$, otherwise $H_i(v/m:p)$ is a list $S$ of tree-edit operations taken from $Q_i$ whose target is $v/m:p$.

We assume that the clients observe the following protocol. 
In particular user $u_i$ can do the following:
\begin{itemize}
\item Perform a valid operation $\tau$ on their local copy of the tree to
produce a new (or extended) node $n$ with nodeid $u/n:q$,
after which they set $H_i(u/n:q)$ to $n$ and they also add $\tau$ to $Q_j$ for
all $j\ne i$.
\item Take an operation $\tau$ from $Q_i$. If the target of $\tau$ is
in $T_i$ then $\tau$ is applied to the tree to produce a node $v/m:p$
and the tree-edit operations in $H_i(v/m:p)$, if any, are inserted back into $Q_i$
 and $H_i(v/m:p)$ is set to $n$.
\end{itemize}
Note, this model can easily be extended to allow for late joiners by having users
maintain a history list $Q'_i$ of all operations they have applied to their tree.  Any user $k$ who joins the group notifies all users it has joined and waits for confirmation from all users. Once this is obtained, it 
requests some client $C_j$ to forward its current history list $Q'_k$ and its current queue, 
$Q_k$ into its own edit queue $Q_i$.
This can be an incremental copy at user $j$'s leisure depending
on the workload.  This may require user $k$ to check each new operation
it receives to make sure that it has not already been received and processed,
but this is not too expensive.

@@

\section{From edit-trees to an optimally efficient collaborative editor}
\label{sec:proof}
In this section we will define the MSET data type $M$ which simultaneously maintains
four views of an editing session
\begin{itemize}
\item $M.S$ - the standard string view of a text-editing session, a sequence of characters from a character set $\Sigma$
\item $M.R$ - a revisions view which is a sequence of characters which each have a visibility attribute (visible or hidden), this is similar to the tracking-changes view of many modern text editors
\item $M.E$ - an edit-string view which is a sequence of attributed characters and marker symbols (start markers $<^u_n$ and end markers $>^u_n$)
\item $M.T$ - an edit tree
\end{itemize}
The three string views are all views of the edit tree $M.T$ obtained by traversing the tree in infix order and recording the characters and or symbols of the specified type.

@@

 {{\tt delete($k$)}} 
Lets handle the deletion case first as it is easiest. The first step is to lookup the element $e$ at position $k$ in $M.S$. This can be done in time $O(\log(N))$.
We can then use $e$ to lookup the node $n$ and offset $q$ of $e$ in that node.
This takes time $O(1)$, as the node and offset are fields of $e$. Finally, we lookup the
nodeid $v/m$ and length $p$ of $n$ and generate the operation $treehide(v/m:p,q)$. This operation takes time $O(\log(N))$. In pseudocode, we have
\begin{verbatim}
delete(k)
  e = M.I.lookupS(k);
  v/m:p = e.nodeid;  q = e.offset
  return treehide(v/m:p,q)
\end{verbatim}

@@

\subsubsection{Case 3a: $g$ is a non-marker element}
In this case, $g$ is either $f$ or a hidden character, but in any case, we can insert $\alpha$ between $e$ and $g$. So in time $O(1)$ we lookup the nodeid $v/m$ and
offset $q$ of $g$ and generate the tree edit operation 
$treeinsert(v/m:p,q,u/n,c)$

@@

\begin{verbatim}
treehide(v/m:p,q)
  n = M.lookupNode(v/m:p);
  e = n.elt[q]
  M.I.hide(e)
\end{verbatim}
